### 2010-03-30 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: clean-up some notes.
Desc@ICFP: merge Colors and Spaces.
Some suggestions to fix some typos

Our work is deeply rooted in the realm of dependently-typed functional
languages. In particular, we have developed this technology in our
on-going work on Epigram. However, in this paper,
we strive to present our technology in the broadest setting
possible. To this end, we are going to present a type theory from the
ground-up.

To the reader new to type theory, this first section will provide a
gentle introduction to dependently-typed systems. To the reader
familiar with type theory, she will recognise the basis of her
favourite type theory, be it Agda or Coq for instance. This scene sets,
we have a common ground for discussion.

@@

In the following, we adopt a standard presentation for our type
theory. It is defined by three forms of judgements: context validity
judgements, typing judgements, and equality judgements. Formally,
these judgements are presented as follow:

@@

Noticeably, we have adopted a judgemental presentation of
equality. Therefore, we are not tied to a particular implementation
choice. In particular, our system has been modelled in Agda, which
features an intensional equality. On the other hand, it is used in
Epigram, which equality has a slightly extensional
flavor~\cite{altenkirch:ott}. We expect users of fully extensional
systems to also find their way through this presentation.

@@

This completes our presentation of the type theory. Assuming a
stratified discipline of universe, the system we have described enjoy
some very strong meta-theoretical properties. Unlike simply typed
languages, such as Haskell, dependently-typed systems are
\emph{strongly normalising}: every program that type-checks
terminates. Moreover, type-checking is decidable and can therefore be
implemented by a terminating algorithm.

\note{Need some care here. Expansion would be good too. I wanted to
  carry the intuition that we are not the bad guys with a
  non-terminating type-checker.}

@@

In the simply-typed world, we are used to see types as a coding
presentation -- presentation of invariants, presentation of
properties. In the dependently-typed world, we also learn to use types
as a coding representation: finite sets being good citizens, they
ought to be democratically represented at the type level. As we will
see, this gives us the ability to name and manipulate them (this is
were the democracy analogy goes crazy, I think).

\note{Did I got the coding presentation vs. coding representation
  story right? No.}

@@

The specification of finite sets is presented in
Figure~\ref{fig:typing-finite-set}. It is composed of three
components. First, we define tags as inhabitants of the $\UId$ type. A
tag is solely an informative token, used for diagnostic
purposes. Finite sets inhabits the $\EnumU$ type. Unfolding the
definition, we get that a finite set is a list of tags. Finally,
elements of a finite set $\V{u}$ belong to the corresponding $\EnumT{\V{u}}$
type. Intuitively, it corresponds to an index -- a number -- pointing
to an element of $u$.

@@

A closer look to its type should reveal a certain familiarity with its
equivalent for sets, $\Pi$. Indeed, $\spi{\!}{\!}$ defines a
\emph{finite function space}, mapping elements of finite sets to
objects in $\Set$. Intuitively, this operator build a right-nested
tuple, effectively mapping an object of $P\:i$ for each $i$ the finite
domain.

@@

However, our situation is significantly different from the one we
faced with encoded data: while we were suffering from a crucial lack
of information, we are now facing too much information, hence loosing
focus. This is a general issue with the usage of codes, as they convey
much more information than what the developer is willing to see.

As we will see in the following section, there exists a cure to this
problem. In a nutshell, instead of being overflown by typing
information, we will put it at work, automatically. The consequence is
that, in such system, working with codes is \emph{practical}: one
should not be worried by information overload, but how to use it as
much as possible. Therefore, we should not be afraid by using code for
practical purposes.

@@

We adapt and formalise this \emph{type propagation} mechanism to our
richer type theory. First of all, we define a high-level syntax,
composed of \emph{expressions}. Expressions are divided into two
syntactic categories: $\exprIn$ into which types are pushed in, and
$\exprEx$ from which types are pulled out. Following the bidirectional
presentation, $\exprIn$ are subject to type \emph{checking}, while
$\exprEx$ are subject to type \emph{synthesis}.

\note{Why don't we need a context validity judgement, again?}

The purpose of type propagation is to \emph{elaborate} an high-level
expression into a term of the type theory, together with its
type. This presentation builds on the experience gained in
implementing Epigram 1~\cite{mcbride:view-from-the-left}. Being
entirely syntax driven, we present our formalism as a system of
syntax-directed inference rules. As the presentation largely mirrors
the inference rules of the type theory, we will ignore the judgements
that are identical. We refer our reader to the associated technical
report~\cite{chapman:desc-tech-report} for the complete system of
rules.

@@

\note{This might deserve a more thorough explanation, with snippet of
  code.}

Finally, and more importantly, we are able to present the developer an
abstraction of \emph{finite} function space. This work as follow. When
an expression $\V{t}$ is pushed into a function which domain is a finite
set, we instead elaborate it as a term in a small $\pi$-type. That is,
a finite function. To realise our promise of a big $\Pi$-type, we
elaborate a call to the $\F{switch}$ eliminator.

The abstraction we have developed in this section are entirely
transparent to the user. In the case of finite function space, the
developer has never to explicitly \emph{call} the $\F{switch}$
eliminator. For diagnosis purposes, a similar, reverse mechanism turns
low-level terms into high-level expressions. Hence, this system, which
complete presentation is out of the scope of this paper, seals the
low-level type theory behind the high-level expressions.

@@

This example teaches us that we should not be afraid by codes, as soon
as type information is available. We have shown how to rationalise
this information in a formal presentation. Hence, we have shown that
programming with such objects is practical. As we introduce more codes
in our theory, we will show how to extend the framework we have
developed so far.

>>>>

Our work is deeply rooted in the realm of dependently-typed functional
languages. In particular, we have developed this technology in our
on-going work on Epigram. However, in this paper,
we strive to present our technology in the broadest setting
possible. To this end, we are going to present a type theory from the
ground up.

To the reader new to type theory, this first section will provide a
gentle introduction to dependently-typed systems. To the reader
familiar with type theory, she will recognise the basis of her
favourite type theory, be it Agda or Coq, for instance. This scene sets,
we have a common ground for discussion.

@@

In the following, we adopt a standard presentation for our type
theory. It is defined by three forms of judgements: context validity
judgements, typing judgements, and equality judgements. Formally,
these judgements are presented as follows:

@@

Noticeably, we have adopted a judgemental presentation of
equality. Therefore, we are not tied to a particular implementation
choice. In particular, our system has been modelled in Agda, which
features an intensional equality. On the other hand, it is used in
Epigram, whose equality has a slightly extensional
flavor~\cite{altenkirch:ott}. We expect users of fully extensional
systems to also find their way through this presentation.

@@

This completes our presentation of the type theory. Assuming a
stratified discipline of universe, the system we have described enjoy
some very strong meta-theoretical properties. Unlike simply typed
languages, such as Haskell, dependently-typed systems are
\emph{strongly normalising}: every program that type-checks
terminates. Moreover, type-checking is decidable and can therefore be
implemented by a terminating algorithm.
\note{Need some care here. Expansion would be good too. I wanted to
  carry the intuition that we are not the bad guys with a
  non-terminating type-checker.}

@@

In the simply-typed world, we are used to see types as a coding
presentation -- presentation of invariants, presentation of
properties. In the dependently-typed world, we also learn to use types
as a coding representation: finite sets being good citizens, they
ought to be democratically represented at the type level. As we will
see, this gives us the ability to name and manipulate them (this is
were the democracy analogy goes crazy, I think).
\note{Did I got the coding presentation vs. coding representation
  story right? No.}

@@

The specification of finite sets is presented in
Figure~\ref{fig:typing-finite-set}. It is composed of three
components. First, we define tags as inhabitants of the $\UId$ type. A
tag is solely an informative token, used for diagnostic
purposes. Finite sets inhabits the $\EnumU$ type. Unfolding the
definition, we get that a finite set is a list of tags. Finally,
elements of a finite set $\V{u}$ belong to the corresponding $\EnumT{\V{u}}$
type. Intuitively, it corresponds to an index -- a number -- pointing
to an element of $\V{u}$.

@@

A closer look to its type should reveal a certain familiarity with its
equivalent for sets, $\Pi$. Indeed, $\spi{\!}{\!}$ defines a
\emph{finite function space}, mapping elements of finite sets to
objects in $\Set$. Intuitively, this operator builds a right-nested
tuple, effectively mapping an object of $P\:i$ for each $i$ the finite
domain.

@@

However, our situation is significantly different from the one we
faced with encoded data: while we were suffering from a crucial lack
of information, we are now facing too much information, hence losing
focus. This is a general issue with the usage of codes, as they convey
much more information than what the developer is willing to see.

As we will see in the following section, there exists a cure to this
problem. In a nutshell, instead of being overflown by typing
information, we will put it at work, automatically. The consequence is
that, in such system, working with codes is \emph{practical}: one
should not be worried by information overload, but how to use it as
much as possible. Therefore, we should not be afraid of using codes for
practical purposes.

@@

We adapt and formalise this \emph{type propagation} mechanism to our
richer type theory. First of all, we define a high-level syntax,
composed of \emph{expressions}. Expressions are divided into two
syntactic categories: $\exprIn$ into which types are pushed in, and
$\exprEx$ from which types are pulled out. Following the bidirectional
presentation, $\exprIn$ are subject to type \emph{checking}, while
$\exprEx$ are subject to type \emph{synthesis}.
\note{Why don't we need a context validity judgement, again?}

The purpose of type propagation is to \emph{elaborate} a high-level
expression into a term of the type theory, together with its
type. This presentation builds on the experience gained in
implementing Epigram 1~\cite{mcbride:view-from-the-left}. Being
entirely syntax driven, we present our formalism as a system of
syntax-directed inference rules. As the presentation largely mirrors
the inference rules of the type theory, we will ignore the judgements
that are identical. We refer our reader to the associated technical
report~\cite{chapman:desc-tech-report} for the complete system of
rules.

@@

Finally, and more importantly, we are able to present the developer an
abstraction of \emph{finite} function space. This works as follow. When
an expression $\V{t}$ is pushed into a function which domain is a finite
set, we instead elaborate it as a term in a small $\pi$-type. That is,
a finite function. To realise our promise of a big $\Pi$-type, we
elaborate a call to the $\F{switch}$ eliminator.

The abstraction we have developed in this section is entirely
transparent to the user. In the case of finite function space, the
developer never has to explicitly \emph{call} the $\F{switch}$
eliminator. For diagnosis purposes, a similar, reverse mechanism turns
low-level terms into high-level expressions. Hence, this system, whose
complete presentation is out of the scope of this paper, seals the
low-level type theory behind the high-level expressions.

@@

This example teaches us that we should not be afraid of codes, as soon
as type information is available. We have shown how to rationalise
this information in a formal presentation. Hence, we have shown that
programming with such objects is practical. As we introduce more codes
in our theory, we will show how to extend the framework we have
developed so far.
